The role of aneuploidy in the evolution of cancer drug resistance

### 2023-07-17 EvolutionaryRescue1.tex
revise with RS and DBW

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Survival probability}

We first assume that aneuploidy provides partial resistance to drug therapy, $\lambda_a>\mu_a$, and that this resistance is significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
We thus rewrite \cref{survival_prob} as
\begin{align*}
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
\text{and} \\
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) . 
\end{align*}
Using the quadratic Taylor expansion $\sqrt{1+x}=1+x/2+O(x^2)$ and assuming $u,v \ll 1$,
we obtain the following approximation for the survival probability of a population initially consisting of a single wildtype cell,
\begin{align}\label{survprobwapprox1}
p_w 
&\approx -\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
\nonumber
%&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\
\end{align}

We now assume that aneuploidy provides tolerance to drug therapy, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$. We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
We rewrite \cref{survival_prob} as
\begin{equation}
\begin{aligned}
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
\text{and} \\
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) .
\end{aligned}
\end{equation}
Since $u,v\ll1$, the term in the root can be approximated using a 1st-order Taylor expansion. So, substituting the expressions for $p_a$ and $p_m$, we have
\begin{equation}\label{survprobwinitial}
\begin{aligned}
p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ %\label{survprobw2}
&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right]\\
%&=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right) .
\end{aligned}
\end{equation}

We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\lambda_a-\mu_a-v\right)^2 < 4\lambda_avp_m$.
We rewrite \cref{survival_prob} as
\begin{equation}
p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a} .
\end{equation}
Using a following Taylor series expansion
\begin{equation*}
\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots,
\end{equation*}
we obtain the approximation
\begin{equation}
\begin{aligned}
p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
&=\frac{\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)^2+4\lambda_avp_m}{8\lambda_a\sqrt{\lambda_avp_m}}\\
&=\frac{4\lambda_avp_m+4\lambda_avp_m\left(1+\frac{\lambda_a-\mu_a-v}{2\sqrt{\lambda_avp_m}}\right)^2}{8\lambda_a\sqrt{\lambda_avp_m}}\\
&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right).
\end{aligned}
\end{equation}
Plugging this in \cref{survprobwinitial}, the survival probability of a population starting from one wildtype individual is
\begin{equation}\label{scenario3}
\begin{aligned}
p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right].
%&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
\end{aligned}
\end{equation}

@@

To estimate the rescue probability $p_{rescue}$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
Thus, the rescue probability is given by 
\begin{subequations}
\begin{align}
p_{rescue} &= 1-\left(1-p_w\right)^N \label{rescue_prob} \\
& \approx 1-\e^{-Np_w} , \label{rescue_prob2}
\end{align}
\end{subequations}
where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.

Applying the approximations for the survival probability $p_w$ from \cref{survprobwapprox1,survprobwinitial,scenario3} in \cref{rescue_prob2} and substituting $\Delta_k=\lambda_k-\mu_k$, we find that the rescue probability can be approximated by
\begin{equation}\label{rescue_prob_approx}
\begin{aligned}
&p_{rescue} \approx \\
  &\begin{cases}
  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
  \end{cases}
\end{aligned}
\end{equation}

% TODO do we need this?
%Note that if instead \cref{survprobwapprox1} we apply \cref{survprobwapproxcorrected}, we have
%\begin{align}\label{survprobresistantaneuploidy}
%p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
%\end{align}
%However, both approximations seem to work very well (\Cref{compare_1st_2nd_order_approx}).

We validate these approximations by comparing them to results of stochastic evolutionary simulations. We find that the approximations work very well (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).

@@

\begin{figure}
% small population size
% switch from tolerance to non-growing
% TODO why no switch in panel B?
% TODO more simulations for 0 > Delta_a > Delta^*_a
% TODO simulations in A?
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
\end{subfigure}
\caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf{A}: $N=10^4$; \textbf{B}: $N=10^8$). 
Markers for simulation results with 95\% CI; 
solid lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob});
dashed lines for the approximate formula (\cref{rescue_prob_approx}).
The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}$.
Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
}
\label{rescue_prob_an_growth}
\end{figure}

@@

%\appendix
%\begin{appendices}
%\renewcommand{\theequation}{\thesection\arabic{equation}}
%\counterwithin*{equation}{section}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section*{Gillespie algorithm}
%\begin{subequations}
%\begin{flalign}
%(+1,0,0)&:\quad \lambda_ww_t,\\
%(-1,0,0)&:\quad \mu_ww_t,\\
%(-1,+1,0)&:\quad uw_t,\\
%(-1,0,+1)&:\quad vw_t,\\
%(0,+1,0)&:\quad \lambda_aa_t,\\
%(0,-1,0)&:\quad \mu_aa_t,\\
%(0,-1,+1)&:\quad va_t,\\
%(0,0,+1)&:\quad \lambda_am_t,\\
%(0,0,-1)&:\quad \mu_am_t.
%\end{flalign}
%\end{subequations}
%\section*{Mean time}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%We define the following events
%\begin{align*}
%\text{EvR}&=\text{``evolutionary rescue occurs''},\\
%\text{EvRA}&=\text{``evolutionary rescue occurs through aneuploidy''},\\
%\text{EvRM}&=\text{``evolutionary rescue occurs through mutation''}.
%\end{align*}
%As a result, the mean time to appearence of the first mutant cell which rescues the population given evolutionary rescue can be expanded as:
%\begin{align*}
%\mathbb{E}\left[\tau|\text{EvR}\right]&=\mathbb{E}\left[\tau|\text{EvR},\text{EvRA}\cap\text{EvRM}^{C}\right]p\left(\text{EvRA}\cap\text{EvRM}^{C}|\text{EvR}\right)\\
%&+\mathbb{E}\left[\tau|\text{EvR},\text{EvRM}\cap\text{EvRA}^{C}\right]p\left(\text{EvRM}\cap\text{EvRA}^{C}|\text{EvR}\right)\\
%&+\mathbb{E}\left[\tau|\text{EvR},\text{EvRM}\cap\text{EvRA}\right]p\left(\text{EvRM}\cap\text{EvRA}|\text{EvR}\right).
%\end{align*}
%The joint probabilities in the previous lines can be written as:
%\begin{align*}
%p\left(\text{EvRA}\cap\text{EvRM}^{C}|\text{EvR}\right)&=\frac{p\left(\text{EvRA}\cap\text{EvRM}^{C}\right)}{p\left(\text{EvR}\right)}
%\approx\frac{p\left(\text{EvRA}\right)\left[1-p\left(\text{EvRM}\right)\right]}{p\left(\text{EvR}\right)},\\
%p\left(\text{EvRM}\cap\text{EvRA}^{C}|\text{EvR}\right)&=\frac{p\left(\text{EvRM}\cap\text{EvRA}^{C}\right)}{p\left(\text{EvR}\right)}
%\approx\frac{p\left(\text{EvRM}\right)\left[1-p\left(\text{EvRA}\right)\right]}{p\left(\text{EvR}\right)},\\
%p\left(\text{EvRM}\cap\text{EvRA}|\text{EvR}\right)&=\frac{p\left(\text{EvRM}\cap\text{EvRA}\right)}{p\left(\text{EvR}\right)}\approx\frac{p\left(\text{EvRM}\right)p\left(\text{EvRA}\right)}{p\left(\text{EvR}\right)}\ll1.
%\end{align*}
%As a result, we have:
%\begin{align*}
%\mathbb{E}\left[\tau|\text{EvR}\right]&\approx\mathbb{E}\left[\tau|\text{EvR},\text{EvRA}\cap\text{EvRM}^{C}\right]p\left(\text{EvRA}\cap\text{EvRM}^{C}|\text{EvR}\right)\\
%&+\mathbb{E}\left[\tau|\text{EvR},\text{EvRM}\cap\text{EvRA}^{C}\right]p\left(\text{EvRM}\cap\text{EvRA}^{C}|\text{EvR}\right).
%\end{align*}
%%%
%\section*{Diffusion approximation}\label{AppendixDiffApprox}
%
%An alternative method to obtain the probability of evolutionary rescue is to utilize a Feller diffusion approximation which is governed by two parameters: the growth rate $r$ and the reproductive variance $\sigma$. The two parameters are obtained from the underlying demographic process as the infinitesimal relative change in mean and variance of $n_t$ over and an infinitesimally small time interval $\Delta t$:
%\begin{align*}
%r&=\lim_{\Delta t\rightarrow0}\frac{\mathbb{E}\left(\Delta n_t|n_t\right)}{\Delta_t n_t},\\
%\sigma&=\lim_{\Delta t\rightarrow0}\frac{\mathbb{V}ar\left(\Delta n_t|n_t\right)}{\Delta_t n_t}.
%\end{align*}
%The rate at which mutants are generated directly from the wildtype is: 
%\begin{align}
%\theta_1=v\bar{\pi}_f\frac{N}{|r_w|},
%\end{align}
%where
%\begin{align}
%\bar{\pi}_f=\int_{0}^\infty\int_{0}^\infty \left(1-\e^{-\frac{2r}{\sigma}}\right) f_r\left(r,\sigma\right)\d r\d\sigma.
%\end{align}
%Letting $f_r\left(r,\sigma\right)=\delta\left(r-r_m\right)\delta\left(\sigma-\sigma_m\right)$ then:
%\begin{align}
%\bar{\pi}_f=1-\e^{-\frac{2r_m}{\sigma_m}},
%\end{align}
%and, as a result, we have:
%\begin{align}
%\theta_1=v\left(1-\e^{-\frac{2r_m}{\sigma_m}}\right)\frac{N}{|\Delta_w|}.
%\end{align}
%The rate at which mutants are generated indirectly from the wildtype through aneuploidy is: 
%\begin{align*}
%\theta_2&=\frac{uN}{|r_w|}\int_{0}^\infty\int_{0}^\infty \left(1-\pi_f\left(r,q\right)\right) f_r\left(r,\sigma\right) p_1^*\left(r,\sigma\right)\d r\d\sigma\\
%&=\frac{uN}{|\Delta_w|}p_1^*\left(r_a,\sigma_a\right)\\
%&=\frac{uN}{|\Delta_w|}\left[1-\exp\left(-\frac{|r_a|}{\sigma_a^2}\left(\sqrt{1+\frac{2\sigma_a^2}{r_a^2}u^*}-1\right)\right)\right]\\
%&=\frac{uN}{|\Delta_w|}\left[1-\exp\left(-\frac{|\Delta_a|}{(\lambda_a+\mu_a)^2}\left(\sqrt{1+\frac{2(\lambda_a+\mu_a)^2}{\Delta_a^2}u^*}-1\right)\right)\right],
%\end{align*}
%where
%\begin{equation}
%u^*=v\left(1-\e^{-\frac{2r_m}{\sigma_m}}\right).
%\end{equation}
%The number of rescue mutations has a Poisson distribution with rate $\theta_1+\theta_2$. As a result, the probability of evolutionary rescue is given by:
%\begin{equation}\label{FellerApprox}
%p_{rescue}=1-\e^{-\left(\theta_1+\theta_2\right)},
%\end{equation}
%which we plot in Figure \ref{P_est_diffusion} as a function of $\Delta_w$.

%\end{appendices}

>>>>

% Daniel commands

\newcommand{\daniel}[1]{\textcolor{blue}{#1}}
\newcommand{\presc}{p_\text{rescue}}
\newcommand{\psgv}{p_\text{SGV}}

@@

@@

To estimate the rescue probability $\presc$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
Thus, the rescue probability is given by 
\begin{subequations}
\begin{align}
\presc &= 1-\left(1-p_w\right)^N \label{rescue_prob} \\
& \approx 1-\e^{-Np_w} , \label{rescue_prob2}
\end{align}
\end{subequations}
where $p_w$ is the probability that the lineage of a single wild-type cell avoids extinction, and the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not necessarily $N p_w$) is small.

The probability for evolutionary rescue $\presc$  (\cref{rescue_prob2}) is $\presc \approx N p_w$ when $N<1/p_w$ and $\presc \approx 1$ when $N > 1/p_w$. Therefore, $N^*=1/p_w$ is the threshold tumor size for rescue above which rescue is very likely (\Cref{rescue_prob_N}). % TODO add 1/p_w to figure as vertical line
Therefore, we can write
\begin{equation}
\presc = 1-e^{-N/N^*} .
\end{equation}
In the Appendix below, we use the theory of multi-type branching processes to find approximate expressions \cref{survprobwapprox1,survprobwinitial,scenario3} for $p_w$ in different regimes. 
Substituting these  into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
\\

In the following, an important quantity is $T^* = (4 v \lambda_a \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
First, if aneuploidy is very rare ($u T^*< 1$), or if aneuploidy is rare ($u < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then rescue will likely occur by a direct resistance mutation in a sensitive cell, such that 
\begin{equation}
N_m^* \approx \frac{\abs{\Delta_w}}{v} \cdot \frac{\lambda_m}{\Delta_m} .
\end{equation}
Otherwise, aneuploidy is frequent enough ($u > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of the tumor. 
The threshold tumor size, $N^*$, can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the log of the aneuploid population size during the critical time,
\begin{equation}\label{rescue_prob_approx}
\begin{aligned}
N_a^* \approx 
% TODO replace XXX with approximations
  \frac{\abs{\Delta_w}}{u} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v} \cdot \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 ,\\ 
  %\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{aligned}
\end{equation}
%The second case has a finer approximation by $N^*_a \approx \frac{\abs{\Delta_w}}{u} \cdot \frac{2\lambda_a}{\Delta_a + 1/T^*}$.
Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).

% TODO move this to appendix
%Substituting these  into \cref{rescue_prob2}, we find that the rescue probability can be approximated by
%\begin{equation}\label{rescue_prob_approx}
%\begin{aligned}
%&\presc \approx \\
%  &\begin{cases}
%  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
%  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
%   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
%   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
%   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
%   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
%  \end{cases}
%\end{aligned}
%\end{equation}

@@

\begin{figure}
% small population size
% switch from tolerance to non-growing
% TODO why no switch in panel B?
% TODO more simulations for 0 > Delta_a > Delta^*_a
% TODO simulations in A?
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
\end{subfigure}
\caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf{A}: $N=10^4$; \textbf{B}: $N=10^8$). 
Markers for simulation results with 95\% CI; 
solid lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob});
dashed lines for the approximate formula (\cref{rescue_prob_approx}).
The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $1/T^*$. % TODO ref to eq
Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
}
\label{rescue_prob_an_growth}
\end{figure}

@@

\appendix
\begin{appendices}
\renewcommand{\theequation}{\thesection\arabic{equation}}
\counterwithin*{equation}{section}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Survival probability of a single lineage}

We first assume that aneuploidy provides partial resistance to drug therapy, $\lambda_a>\mu_a$, and that this resistance is significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
We thus rewrite \cref{survival_prob} as
\begin{align*}
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
\text{and} \\
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) . 
\end{align*}
Using the quadratic Taylor expansion $\sqrt{1+x}=1+x/2+O(x^2)$ and assuming $u,v \ll 1$,
we obtain the following approximation for the survival probability of a population initially consisting of a single wildtype cell,
\begin{align}\label{survprobwapprox1}
p_w 
&\approx -\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
\nonumber
%&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\
\end{align}
Now $u v$ is very small, and if we assume $v \ll u$, we have
\begin{equation}
p_w \approx \frac{u}{\abs{\Delta_w}} \cdot \frac{\Delta_a}{\lambda_a} .
\end{equation}

We now assume that aneuploidy provides tolerance to drug therapy, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$. We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
We rewrite \cref{survival_prob} as
\begin{equation}
\begin{aligned}
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
\text{and} \\
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) .
\end{aligned}
\end{equation}
Since $u,v\ll1$, the term in the root can be approximated using a 1st-order Taylor expansion. So, substituting the expressions for $p_a$ and $p_m$, we have
\begin{equation}\label{survprobwinitial}
\begin{aligned}
p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ %\label{survprobw2}
&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right] \\
&=\frac{v\Delta_m}{\lambda_m \abs{\Delta_w}}\left(\frac{u}{\abs{\Delta_a}}+1\right) .
\end{aligned}
\end{equation}

We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\lambda_a-\mu_a-v\right)^2 < 4\lambda_avp_m$.
We rewrite \cref{survival_prob} as
\begin{equation}
p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a} .
\end{equation}
Using a following Taylor series expansion for small $\left(\lambda_a-\mu_a-v\right)^2 / 4\lambda_avp_m$,
\begin{equation*}
\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots,
\end{equation*}
we obtain the approximation
\begin{equation}
\begin{aligned}
p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
&=\frac{\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)^2+4\lambda_avp_m}{8\lambda_a\sqrt{\lambda_avp_m}}\\
&=\frac{4\lambda_avp_m+4\lambda_avp_m\left(1+\frac{\lambda_a-\mu_a-v}{2\sqrt{\lambda_avp_m}}\right)^2}{8\lambda_a\sqrt{\lambda_avp_m}}\\
&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right).
\end{aligned}
\end{equation}
Plugging this in \cref{survprobwinitial}, the survival probability of a population starting from one wildtype individual is
\begin{equation}\label{scenario3}
\begin{aligned}
p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right].
%&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
\end{aligned}
\end{equation}

\end{appendices}
